Splints of root systems for special Lie subalgebras

### 2015-03-21 AffineLieAlgebras/Splints/special-paper.tex
Submitted to TMPh on February 2015

\documentclass[12pt]{iopart}

%\usepackage{amsfonts}
\usepackage{pb-diagram}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\usepackage{iopams,setstack}
\usepackage{multicol}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[english]{babel}

@@

\ftc{Splints of root systems for special Lie subalgebras}

\author{V.D.~Lyakhovsky$^1$, A.A.~Nazarov$^{1,2}$, P.I.~Kakin$^{1}$}
\address{ $^1$ Department of High-energy and elementary particle physics,  St Petersburg State University, 198904, Saint-Petersburg, Russia}
\ead{ $^{2}$ e-mail: antonnaz@gmail.com}

@@

\submitto{\jpa}

@@

Assume that the root system $\Delta_{\gf}$ splinters as $\Delta_{\gf}=\Delta_{\af} \cup
\phi(\Delta_{\sfr})$, where $\phi$ is an embedding of a root system $\Delta_{\sfr}$ of some semisimple
Lie algebra $\sfr$. Then branching coefficients $b^{(\mu)}_{\nu}$ for the reduction
$L^{(\mu)}_{\gf\downarrow \af}$ coincide with the weight multiplicities $m^{(\tilde \mu)}_{\nu}$ in the $\sfr$-representations provided certain technical condition holds \cite{2011arXiv1111.6787L}.
The highest weight $\tilde\mu$ of the $\sfr$-representation is calculated from the Dynkin labels of the highest
weight $\mu$ of the $\gf$-module.

In the next section \ref{sec:splints-root-systems} we review the classification of splints for simple
root systems and the results for branching coefficients. Then we move to the study of special embeddings
which is the main subject of the present paper. We consider special embeddings of Lie subalgebras
into a Lie algebra. In this case the root system of the subalgebra is not contained in the root system of
the algebra. So the original motivation for splints is not applicable. But we can consider the
projection of the root system of the algebra on the root space of the subalgebra. Such a projection is not a
root system anymore, but it satisfies milder conditions (Section \ref{sec:spec-embedd-proj}). It is
possible to classify the splints of the projection using the fact that the adjoint representation of the algebra is decomposed into the direct sum of representations of the subalgebra. We show that if there is a splint then branching coefficients coincide with weight multiplicities of representations of another Lie algebra (Section \ref{sec:splints-spec-embedd}). Thus in this case branching coefficients are very easy to compute.

@@

\section{Splints of root systems and regular subalgebras}
\label{sec:splints-root-systems}
The notion and the classification of
splints for simple root systems were introduced in the paper \cite{richter2008splints}.

Consider a simple Lie algebra $\mathfrak{g}$ and its regular subalgebra $%
\mathfrak{a}\hookrightarrow \mathfrak{g}$ such that $\mathfrak{a}$
is a
reductive subalgebra $\mathfrak{a \subset g}$ with correlated root spaces: $%
\mathfrak{h}_{\mathfrak{a}}^{\ast }\subset \mathfrak{h}_{\mathfrak{g }%
}^{\ast }$. Let $\mathfrak{a}^{\mathfrak{s}}$ be a semisimple summand of
$\mathfrak{a}$,
this means that $\mathfrak{a}=\mathfrak{a}^{\mathfrak{s}} \oplus \mathfrak{u}(1)\oplus %
\mathfrak{u}(1)\oplus \dots$. We shall consider $\mathfrak{a}^{\mathfrak{s}}$
to be a proper regular subalgebra and $\mathfrak{a}$ to be the
maximal subalgebra with $\mathfrak{a}^{\mathfrak{s}}$ fixed, so that the rank
$r$ of $\af$ is equal to that of $\mathfrak{g}$.

The computation of branching coefficients relies on the roots $\Delta_{\gf}\setminus \Delta_{\af}$
\cite{2010arXiv1007.0318L} so splint  is naturally connected to branching coefficients for regular
subalgebras.

Only three types of splints are
injective and thus are naturally connected to branching \cite{richter2008splints}:
\begin{equation}
\label{tab:1}
\begin{array}{cc||c|c}
\hbox{type} & \hspace{0.25in}\Delta \hspace{0.25in} & \hspace{0.25in}\Delta
_{\frak{a}}\hspace{0.25in} & \hspace{0.25in}\Delta _{\sfr}\hspace{0.25in}
\\ \hline\hline
\hbox{(i)} & G_{2} & A_{2} & A_{2} \\
& F_{4} & D_{4} & D_{4} \\ \hline
\hbox{(ii)} & B_{r}(r\geq 2) & D_{r} & \oplus ^{r}A_{1} \\
(*)& C_{r}(r\geq 3) & \oplus ^{r}A_{1} &  D_{r} \\ \hline
\hbox{(iii)} & A_{r}(r\geq 2) & A_{r-1}\oplus u\left( 1\right)  & \oplus
^{r}A_{1} \\
& B_{2} & A_{1}\oplus u\left( 1\right)  & A_{2}
\end{array}
\end{equation}

Each row in the table gives a splint $(\Delta _{\frak{a}},\Delta _{\sfr})
$ of the simple root system $\Delta $. In the first two types both $\Delta _{%
\frak{a}}$ and $\Delta _{\sfr}$ are embedded metrically. Stems in the
first type splints are equivalent and in the second are not. In the third
type splints only $\Delta _{\frak{a}}$ is embedded metrically. The summands $%
u\left( 1\right) $ are added to keep $r_{\frak{a}}=r$. This does not change
the principle properties of branching but makes it possible to use the
multiplicities of $\sfr$ -modules without further projecting their
weights.

Note that in the case of $C_{r}$-series (marked with the star in table \ref{tab:1}) the metrical
embedding $\Delta_{D_{r}}\to \Delta_{C_{r}}$ does not lead to the appearance of the regular subalgebra
$\mathfrak{a}$ (See \cite{dynkin1952semisimple}).

In the paper \cite{2011arXiv1102.1702L} it was shown that

\begin{Prop}
There is a decomposition of a singular element of the algebra $\gf$ into singular elements of the subalgebra $\sfr$:
\begin{equation}
 \Phi ^{\mu }_{\gf} = \sum_{w\in W_{\frak{a}}}\epsilon\left( w\right)w\left( e^{\mu +\rho _{\gf}}\phi\left( e^{-\widetilde{\mu }}\Psi^{\widetilde{\mu }}_{\sfr}\right)\right),
\end{equation}
where $\epsilon\left( w\right)$ is a determinant of the element $w$ of the Weyl group $W_{\af}$ of
the subalgebra $\af$. The action of the Weyl group on the weights is extended to the algebra of formal
exponents by the rule $w\left(e^{\nu}\right)=e^{w\nu}, w\in W$. Hence, the multiplicity $M_{\left(
      \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}$ of the weight $\widetilde{\nu }$ from the
  weight diagram of the algebra $\sfr$ with the highest weight $\tilde\mu$ defines the branching
  coefficient $b_{\nu }^{(\mu )}$ for the highest weight $\nu =\left( \mu -\phi \left(
      \widetilde{\mu }-\widetilde{\nu }\right) \right) $:
\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{(\mu )}=M_{\left( \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}. 
\label{bran1}
\end{equation}
\end{Prop} 
%% Rewrite this property informally

@@

As an example we present a projection $\Delta'$ of the root system of $D_{4}$ to the root system of the special
subalgebra $A_{2}$ (Fig. \ref{fig:d4-a2_splint}).

\begin{figure}[h!bt]
  \noindent\centering{
   \includegraphics[width=40mm]{Special-A2-D4-all-2}
  }
  \caption{Projection of $D_{4}$ ($so(8)$)-root system onto the root space of the special subalgebra $A_{2}$
    ($su(3)$). Note that this root system is $G_{2}$ but with non-trivial multiplicities. }
 \label{fig:d4-a2_splint}
\end{figure}

@@

Nevertheless, the union of diagrams of the two fundamental representations of $A_{2}$ with the root
system of $A_{2}$ produces a weight diagram of $G_{2}$. This case corresponds to splint
$\Delta_{G_{2}}=\varphi_{1}( \Delta_{A_{2}})\cup \varphi_{2}(\Delta_{A_{2}})$ which is connected
with the regular subalgebra $A_{2}\subset G_{2}$ (see Section \ref{sec:splints-root-systems}).

@@

\section*{Conclusion}
\label{sec:conclusion}

@@

\bibliography{special-bibliography}{} 
\bibliographystyle{iopart-num}

>>>>

\documentclass[12pt]{article}

\baselineskip=20pt
\hsize=340pt
\vsize=490pt

\usepackage{amssymb}

\usepackage{color,graphicx}

\usepackage{amsmath}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{pb-diagram}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\usepackage{multicol}
\usepackage{color}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage[english]{babel}

@@

\title{Splints of root systems for special Lie subalgebras}

\author{V.D.~Lyakhovsky$^1$, A.A.~Nazarov$^{1,2}$, P.I.~Kakin$^{1}$\\
  {\small $^1$ Department of High-energy and elementary particle physics,}\\ {\small St Petersburg State University}\\
  {\small 198904, Saint-Petersburg, Russia,}\\
  {\small $^{2}$ e-mail: antonnaz@gmail.com}}
%\date{}
%\address{ $^1$ Department of High-energy and elementary particle physics,  St Petersburg State University, 198904, Saint-Petersburg, Russia}
%\address{ $^{2}$ e-mail: anton.nazarov@hep.phys.spbu.ru}

\maketitle

@@

@@

Assume that the root system $\Delta_{\gf}$ splinters as $\Delta_{\gf}=\Delta_{\af} \cup
\phi(\Delta_{\sfr})$, where $\phi$ is an embedding of a root system $\Delta_{\sfr}$ of some semisimple
Lie algebra $\sfr$. Then branching coefficients $b^{(\mu)}_{\nu}$ for the reduction
$L^{(\mu)}_{\gf\downarrow \af}$ coincide with the weight multiplicities $m^{(\tilde \mu)}_{\nu}$ in the $\sfr$-representations provided certain technical condition holds \cite{2011arXiv1111.6787L}.
The highest weight $\tilde\mu$ of the $\sfr$-representation is calculated from the Dynkin labels of the highest
weight $\mu$ of the $\gf$-module.
The proof of the coincidence of the branching coefficients with the weight multiplicities of
$\sfr$-module is based upon the decomposition of singular element of the algebra $\gf$ into singular
elements for $\sfr$:
\begin{equation}
 \Phi ^{\mu }_{\gf} = \sum_{w\in W_{\frak{a}}}\epsilon\left( w\right)w\left( e^{\mu +\rho _{\gf}}\phi\left( e^{-\widetilde{\mu }}\Psi^{\widetilde{\mu }}_{\sfr}\right)\right),
\end{equation}
where $\epsilon\left( w\right)$ is a determinant of the element $w$ of the Weyl group $W_{\af}$ of
the subalgebra $\af$. The action of the Weyl group on the weights is extended to the algebra of formal
exponents by the rule $w\left(e^{\nu}\right)=e^{w\nu}, w\in W$. Hence, the multiplicity $M_{\left(
      \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}$ of the weight $\widetilde{\nu }$ from the
  weight diagram of the algebra $\sfr$ with the highest weight $\tilde\mu$ defines the branching
  coefficient $b_{\nu }^{(\mu )}$ for the highest weight $\nu =\left( \mu -\phi \left(
      \widetilde{\mu }-\widetilde{\nu }\right) \right) $:
\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{(\mu )}=M_{\left( \sfr\right) \widetilde{\nu }}^{\widetilde{\mu }}. 
\label{bran1}
\end{equation}

We use similar approach to study special embeddings. We consider special embeddings of Lie
subalgebras into a Lie algebra. In this case the root system of the subalgebra is not contained in
the root system of the algebra. So the original motivation for splints is not applicable. But we can
consider the projection of the root system of the algebra on the root space of the subalgebra. Such
a projection is not a root system anymore, but it satisfies milder conditions (Section
\ref{sec:spec-embedd-proj}). It is possible to classify the splints of the projection using the fact
that the adjoint representation of the algebra is decomposed into the direct sum of representations
of the subalgebra. We show that if there is a splint then branching coefficients coincide with
weight multiplicities of representations of another Lie algebra (Section
\ref{sec:splints-spec-embedd}). Thus in this case branching coefficients are very easy to compute.

@@

@@

For example the projection of the root system of $D_{4}$($so(8)$) to the root system of the special
subalgebra $A_{2}$($su(3)$) coincides with the root system $G_{2}$, but the short roots have the multiplicity $3$.

@@

Nevertheless, the union of diagrams of the two fundamental representations of $A_{2}$ with the root
system of $A_{2}$ produces a weight diagram of $G_{2}$. This case corresponds to splint
$\Delta_{G_{2}}=\varphi_{1}( \Delta_{A_{2}})\cup \varphi_{2}(\Delta_{A_{2}})$ which is connected
with the regular subalgebra $A_{2}\subset G_{2}$ (see \cite{2011arXiv1111.6787L}).

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\bibliography{special-bibliography}{} 
\bibliographystyle{utphys}
